---
title: "Cultural Competence Matrix: A Visualization Walkthrough"
subtitle: "Public-health access in marginalized populations"
author: "Kamalakanta Gahan"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", dpi = 150
)
```

## Overview

This notebook walks through the **Cultural Competence Matrix** end to end: it
loads the scored data, computes access indices, and renders the four figures.
The framework is population-agnostic — the same code runs on any dataset that
follows the schema in `data/`. Here it is illustrated with the documented
**Malayali (Sitheri Hills)** case from the author's MA dissertation, alongside
illustrative comparison populations.

```{r load}
library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)

source("R/load_data.R")
source("R/scoring.R")
source("R/theme_ccm.R")

meta      <- load_domains()
mat       <- load_matrix(meta = meta)
pluralism <- load_pluralism()

glimpse(mat)
```

## The six domains

Each domain is grounded in health-services / cultural-competence literature and
mapped to Betancourt's tri-level model and Andersen's access stages.

```{r domains}
meta |>
  select(domain_id, domain, betancourt_level, andersen_stage, access_type) |>
  knitr::kable()
```

## Scoring: from cells to access indices

The matrix records an ordinal **0–3** score per (population × domain). We rescale
to a 0–100 index and split **capacity** (provision: D1–D3) from **realized**
(dignified use: D4–D6). Their difference is the **conversion gap**, and a simple
threshold rule assigns a transferable diagnostic label.

```{r indices}
indices <- summarise_access(mat)

indices |>
  select(population, composite_index, capacity_index, realized_index,
         conversion_gap, diagnostic) |>
  arrange(desc(composite_index)) |>
  knitr::kable(digits = 1)
```

A second lens rolls the same scores up to Betancourt's organizational /
structural / clinical levels:

```{r levels}
compute_level_profile(mat) |>
  pivot_wider(names_from = betancourt_level, values_from = level_index) |>
  knitr::kable(digits = 1)
```

## Figure 1 — Scored matrix heatmap

The hero view: six domains across all populations.

```{r fig1}
heat_df <- mat |> mutate(short_label = fct_rev(short_label))

ggplot(heat_df, aes(population, short_label, fill = score)) +
  geom_tile(color = "white", linewidth = 1.1) +
  geom_text(aes(label = score), color = "grey15", fontface = "bold", size = 4) +
  scale_fill_gradient2(low = CCM_GRADIENT_LOW, mid = CCM_GRADIENT_MID,
                       high = CCM_GRADIENT_HIGH, midpoint = 1.5,
                       limits = c(0, 3), breaks = 0:3, name = "Score") +
  scale_x_discrete(position = "top") +
  labs(title = "Cultural Competence Matrix", x = NULL, y = NULL) +
  theme_ccm() +
  theme(axis.text.x = element_text(angle = 20, hjust = 0, face = "bold"),
        panel.grid = element_blank())
```

## Figure 2 — Radar profiles

The *shape* of access. The documented case (structural presence, organizational
and clinical gaps) contrasts with an illustrative profile that inverts it.

```{r fig2, fig.height=7}
radar_pops <- c("Malayali (Sitheri Hills)", "Remote Indigenous community")

radar_closed <- mat |>
  filter(population %in% radar_pops) |>
  select(population, short_label, score) |>
  mutate(short_label = factor(short_label, levels = levels(meta$short_label))) |>
  group_by(population) |>
  group_modify(~ bind_rows(.x, .x[1, ])) |>
  ungroup()

ggplot(radar_closed, aes(short_label, score, group = population,
                         color = population, fill = population)) +
  geom_polygon(alpha = 0.18, linewidth = 1.1) +
  geom_point(size = 2.6) +
  coord_polar() +
  scale_y_continuous(limits = c(0, 3), breaks = 0:3) +
  scale_color_manual(values = unname(CCM_CAT_COLORS), name = NULL) +
  scale_fill_manual(values = unname(CCM_CAT_COLORS), name = NULL) +
  labs(title = "Access profiles", x = NULL, y = NULL) +
  theme_ccm() +
  theme(legend.position = "bottom")
```

## Figure 3 — The conversion gap

Does built capacity become dignified use? Distance between dots = access the
system provides but does not realize (Andersen).

```{r fig3}
gap_df <- indices |>
  mutate(population = fct_reorder(population, composite_index))

gap_long <- gap_df |>
  pivot_longer(c(capacity_index, realized_index),
               names_to = "kind", values_to = "index") |>
  mutate(kind = recode(kind,
                       capacity_index = "Capacity (provision)",
                       realized_index = "Realized (dignified use)"))

ggplot(gap_df, aes(y = population)) +
  geom_segment(aes(x = realized_index, xend = capacity_index, yend = population),
               color = "grey75", linewidth = 1.6) +
  geom_point(data = gap_long, aes(x = index, color = kind), size = 4.2) +
  scale_color_manual(values = c("Capacity (provision)" = "#2166ac",
                                "Realized (dignified use)" = "#e07a5f"),
                     name = NULL) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(title = "The conversion gap", x = "Access index (0-100)", y = NULL) +
  theme_ccm() +
  theme(legend.position = "bottom")
```

## Figure 4 — Medical pluralism in transition

A case-specific slope chart of the dissertation's directional finding:
ethnomedicine contracts as the PHC embeds. Values are a schematic of a
qualitative finding, not measured percentages.

```{r fig4}
plur <- pluralism |>
  mutate(system = factor(system,
                         levels = c("Traditional medicine (Ta-ta)",
                                    "Primary Health Centre (PHC)",
                                    "Community Health Centre (CHC)")))

ggplot(plur, aes(period, reliance, group = system, color = system)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 3.4) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 0.8)) +
  scale_color_manual(values = c("#3d8c5f", "#1b4965", "#e07a5f"), name = NULL) +
  labs(title = "Medical pluralism in transition (Sitheri Hills)",
       x = NULL, y = "Share of care-seeking (illustrative)") +
  theme_ccm() +
  theme(legend.position = "bottom")
```

## Interpretation

Read across the indices and the heatmap together. The documented case shows
relatively strong **structural access** and **trust/pluralism** (the elements the
state has installed) but weak **organizational representation**, **communication
fit**, and **dignity** — the elements that require the system to adapt to the
population. In Betancourt's terms: *structural presence with organizational and
clinical gaps*. That diagnostic — not the specific scores — is the framework's
portable output, and it points a program toward the level at which to intervene.

To apply the framework to a new population, edit only `data/example_matrix.csv`;
every table and figure above regenerates unchanged.

---

*Comparison-population data are synthetic and for illustration only. Construct
definitions are grounded in PubMed-indexed literature; see the README references.*
